Haploid Selection, Sex Ratio Bias, and Transitions Between Sex-Determination Systems

### 2018-05-17 LaTeX/response.tex

\section{Reviewer 1}

The turnover within a species from one genetic means of sex determination to another is surprisingly common, and it poses a challenging puzzle for evolutionary geneticists.  Existing theory on the topic suggests that there must be either sexually-antagonistic selection or selection variants that bring the sex ratio closer to 1:1.  This paper builds explicit models of the situation and asks about the role of selection in the haploid stage (i.e. meiotic drive and gametic selection), and finds that such effects can also serve as the primary driver of turnover of sex determining mechanisms.  While this is a rather focused topic, it is a striking evolutionary puzzle, and the paper presents a very satisfying solution that is much more general than we had previously thought.  It also overturns the primacy of Fisherian sex ratio evolution as the driver of the dynamics of genes of this sort.

The authors devise a satisfyingly simple genetic scenario, with one ancestral sex-determining locus, one locus under selection, and a third locus for the nascent sex determining locus.  They manage to collapse some messy algebra into terms like the ``haplotypic growth rate" that allow the reader to intuit many of the results.  I only have a few suggestions:

\textcolor{blue}{
We thank the reviewer for their careful consideration of our manuscript and pithy summary. We have incorporated the sentiment of ``a striking evolutionary puzzle'' to help frame our study, L51. Below, we itemize further improvements we have made that correspond to more specific reviewer comments. 
}

@@

\noindent\subsection{}
A glitch appears in the section ``Loose linkage with the ancestral sex-determining region" on line 270, where the text reads, ``To leading order in selection, the leading eigenvalues are..."

\noindent\paragraph{Response}
\textcolor{blue}{Thank you. We have dropped the words ``the leading eigenvalues are".}

\noindent\subsection{}
Gametic selection among sperm is a bit different from sperm competition, since the latter involves pair-wise (or multi-way) bouts, whereas gametic selection assigns constant fitnesses to each haplotype. I would replace occurrences of ``sperm competition" with ``male gametic selection", or maybe ``sperm gametic competition". To be consistent, replace ``egg competition" with ``female/egg gametic competition".

\noindent\paragraph{Response}
\textcolor{blue}{We have made the suggested replacements.
In addition, we have added a more detailed discussion of how gametic competition relates to the number of mating partners involved, L108-118 (see also response XXXXXX). 
Specifically, we note that our model explicitly considers within-ejaculate competition or competition among sperm from many males but not competition among sperm from pairs (or small numbers) of males. 
%n the introduction we mention that sperm competition has been shown to alter allele frequencies and fitness, and in this case we now mention this was shown within a single ejaculate (L68).
}

@@

\noindent\subsection{}
Likewise, the results yield three major insights:\\
(1) SD transitions can, counterintuitively, potentially INCREASE sex ratio bias\\
(2) Tight linkage and sexual antagonism are not required for transitions to occur\\
(3) Under haploid selection, conditions for the spread of new sex-determining alleles are extremely permissive.\\
Again, all of these points are there, but they could be better drawn out than they currently are. The results, as written, are hard-going, and I expect will be particularly challenging for readers that are not heavily immersed in evolutionary theory.

\noindent\paragraph{Response}
\textcolor{blue}{
As discussed further below (response XXXXX), we have highlighted conclusions in the results section as suggested. 
To further reduce the technicality of our presentation, we have also revised the results section. 
In particular, we have restructured the presentation of the subsections ``Generic invasion by a neo-Y or neo-W'' and ``Tight linkage with the ancestral sex-determining locus''. 
In the ``Generic invasion by a neo-Y or neo-W" section, we have moved some details to the supplementary material (particularly, proofs regarding the leading eigenvalue, formally LXXX-XXX). 
Now, we focus on explaining the intuition behind our technical results.  
For example, we first discuss the `haplotypic growth rates' and 'dissociative forces' (L158-170 and `interpreting this condition...', L173).
We then lay out three specific conclusions (L179-199). 
In the ``Tight linkage'' section, we now first describe specific cases why neo-W's can have improved female fitness, L235-259, with more focus on non-technical interpretation (e.g., L237-240). 
Then we discuss the fact that looser sex-linkage can evolve (L260-275, Conclusion 2). 
This re-organisation allows us to break down the logic behind Conclusion 2 more clearly. 
}

\noindent\subsection{}
One potential way to help readers -- particularly those interested in what the paper has to say, but are not mathematically inclined -- is to conspicuously highlight each important result and briefly place it into context of earlier theory. But at the same time, the authors have done a great job of drawing out general mathematical results from a relatively complicated model -- their mathematical insights ARE the results and I am not suggesting that they be minimized or shunted into appendices.

@@

One tactic for including analytical details which specialists will want to see, while allowing less mathematically inclined readers to easily find each main result, is to place short and accessible summaries at the close of each section of the Results. Such a tactic has been put to good use in a hugely influential paper by Allen Orr (1998 Evolution), which I expect the authors will be familiar with. Orr presents the major mathematical results of his analysis alongside prominently labelled punchline paragraphs that summarize each of his main results in a concise and easily understandable manner. If the authors choose to take this approach, please be clear to place the take-home messages in context: clearly state whether the result confirms previous theory, or represents a new result or contradiction of an earlier result.

\noindent\paragraph{Response}
\textcolor{blue}{
This is an excellent suggestion and has greatly improved the clarity of presentation in our manuscript. In the results, we have specifically highlighted our main conclusions in the style of Orr 1998. We have chosen slightly different conclusions from the three suggested by the reviewer: 
%We now give 3 main conclusions at the end of the introduction (L69) which echo those mentioned by the reviewer (looser linkage can evolve, equal selection to increase or decrease the sex ratio, ESD does not always invade biased sex ratio; all of which suggest haploid selection makes conditions for transitions more permissive).
\begin{enumerate}
\item[(1)] Selection on loci in or near the non-recombining region around the ancestral sex-determining locus ($r\approx0$) prevents cis-GSD transitions (XY $\leftrightarrow$ XY, ZW $\leftrightarrow$ ZW) but can spur trans-GSD transitions (XY $\leftrightarrow$ ZW). (L228-230)
\item[(2)] With tight linkage between a selected locus and the ancestral sex-determining locus ($r\approx0$), trans-GSD transitions (XY $\leftrightarrow$ ZW) can be favoured by selection \textit{even if they weaken sex-linkage} ($r<R$), potentially shifting sex determination to a different chromosome ($R=1/2$). 
Such transitions can also lead to the maintenance of multifactorial sex-determination systems. (L282-286)
\item[(3A)] New sex-determining alleles (causing cis- \textit{or} trans-GSD transitions) can spread if they arise more closely linked with a locus that experiences selection than the ancestral-sex-determining locus is ($R<r$). (L324-326)
\item[(3B)] New sex-determining alleles (causing trans-GSD transitions, XY $\leftrightarrow$ ZW) can spread even if they arise less closely linked with a locus that experiences selection than the ancestral-sex-determining locus is ($r<R$), requiring haploid selection. (L327-330)
\item[(4)] When selection is weak relative to recombination, the presence of haploid selection equally favors the spread of new sex determination systems that reduce sex-ratio bias (benefiting from Fisherian sex ratio selection) or generate a sex-ratio bias (benefiting from associations with selected alleles). (L350-353)
\item[(5)] Transitions from genetic to environmental sex-determination are not straightforwardly predicted by selection to balance the zygotic sex ratio when haploid selection is present.  (L411-413)
%\vspace{0.25cm}\\
\end{enumerate}
These conclusions are also more clearly laid out in the introduction (L48-49, L74-85) and are referenced directly in summarisations in the discussion (L424-432, L543-554). 
}

\textcolor{blue}{Of these conclusions, (3A) and (3B) have some precendent. (3A) is similar to studies by van Doorn and Kirkpatrick (2007,2010) that do not include haploid selection. These comparisons are made immediately above the conclusions (L316-330) and mentioned elsewhere (L55, L77, L489). 
Including haploid selection, Ubeda et al. (2015) also involved examples similar to (3A), except that the starting state was ESD. 
Kozielska et al. (2010) also had numerical examples that involved trans-GSD transitions with ($R>r$), as stated in (3B), but focussed only on sex ratio selection. 
We discuss how our results compare to these studies in the introduction (L58-75), discussion (L433-447) and results (L193, L363-369).
Nevertheless, we have kept Conclusions 3A and 3B as highlighted conclusions. 
This is partly to highlight these conclusions for clarity of communication and partly because our results are more general. 
That is, as stated by the reviewer, our results ``converge on previously-identified conditions for the spread of new SD alleles, while also broadening the range of scenarios and conditions that can favour transitions between SD systems''}
%As (1A), (1B), and (2) have some antecedents in the literature, we discuss these immediately below each conclusion. 
%We also discuss, immediately below each conclusion, its novelty.
%These are all new results, although (1A) was shown for the special case of $r=0$, $R=1/2$, and ploidally-antagonistic selection in Kozielska et al 2010, which we now mention immediately below (1A).
}

@@

\section{Reviewer 3}

This work (henceforth [SOO]) explores the role of both diploid and haploid selection on the evolution of new sex-chromosomes and or transitions between sex-determination systems. For that purpose, the authors develop a population genetics model of three loci with two alleles in each locus. The first locus X is the ancestral sex-determining locus, the second locus A is the locus under diploid and haploid selection. The third locus M is the novel sex-determining locus.

\noindent\subsection{}
I find the article is not clearly written and could benefit from improving the presentation of the main results. 
%
%\noindent\paragraph{Response}
%\textcolor{blue}{We hope the improvements discussed above (e.g., Response 2.1 and 2.2) and below have increased the clarity of our writing, especially regarding the main results.}
%
%\noindent\subsection{}
I also find the article claims that: (a) the modeling work is novel and (b) the findings are unexpected to be only partially justified.
%
%\noindent\paragraph{Response}
%\textcolor{blue}{As described at length below, we now give more weight to two previous papers (Kozielska et al 2010 and Ubeda et al 2015) and more precisely indicate which of our results are new and unexpected.}
%
%\noindent\subsubsection{}
The works of Kozielska 2014 and Ubeda 2015 model the combined effects of both haploid and diploid selection on the evolution of new sex-chromosomes. The work of [SOO] is thus not novel in considering both haploid and diploid selection on the evolution of new sex-chromosomes. It does generalize previous work by considering sex-specific viability selection and get some nice analytical results.

\noindent\paragraph{Response}
%\textcolor{red}{The abstract and summary still seem to be a little over-selling...}
\textcolor{blue}{
We have performed an extensive re-arrangement to improve the clarity of our writing, especially regarding the main results. 
As described in Response 2.1 (point 6) and below, we now provide a prominent and longer discussion of Kozielska et al 2010 and Ubeda et al 2015 (L57-69) and more precisely indicate which of our model features and results are new and unexpected in light of these previous results (see Response 2.2).
While related, these two studies considered particular cases that we do not consider at length. 
For example, Ubeda et al consider ESD-GSD transitions. 
Nevertheless, these transitions involve increased linkage with a haploid selected locus, which is precedent for our Conclusion 3A (which considers only GSD-GSD transitions).
We now mention this in the Introduction (L59-61) and again in the Discussion (L432).  
Similarly, Kozielska et al involves a special case of GSD transitions during which sex-linkage goes from perfect to none, which is is an  example of our Conclusion 3B.
This is twice mentioned (L64-67, L237). 
Despite this precedent, we find the permissive parameter space allowing GSD-GSD transitions surprising. 
%Especially because their main conclusion involved the strong influence of sex-ratio selection, which is explicitly not a driver during many such transitions. 
In addition, our Conclusions 3A and 3B reconcile Kozielska et al and Ubeda et al, explaining that sex-ratio selection is not an overwhelming force driving transitions in sex determination. 
We believe that our new presentation makes the novel and unexpected features of our results much more apparent. 
%To be clear, we do not claim to be the first to include both haploid and diploid selection to a model investigating the evolution of new sex-chromosomes and we do not consider the addition of sex-specific selection or analytical results for previously numerical models to be our most significant contributions (again, see Responses 2.1 and 2.2 and below for more detail on what these are). 
}

@@

\noindent\subsubsection{}
I will not provide further examples, but it is surprising thatI the work of Ubeda et al 2015, which is closely related to the one presented here, is cited in page 12 supporting a statement only tangential to the work of Ubeda et al 2015.

\noindent\paragraph{Response}
\textcolor{blue}{Ubeda et al 2015 was already cited on line 396 of the original submission, where we mentioned that it showed that sex-ratio biases could become more severe during a transition.
Nevertheless, we now discuss Ubeda et al 2015 in the Introduction, Results, and Discussion (see Responses above).}

\noindent\subsection{}
In terms of modeling the distinction between meiotic drive and gametic competition is unnecessary and only complicates notation. Furthermore, from biological perspective the division seems blurry with molecular biologists considering that meiotic drive only happens in females and never in males and what is often called meiotic drive in males is gamete competition. I would suggest using a single parameter measuring transmission distortion.

\noindent\paragraph{Response}
\textcolor{blue}{We do not agree. 
Given our definition of meiotic drive and gametic competition (L106-116), the two processes are different.
In particular, meiotic drive is assumed to occur during meiosis, and thus is ``gametic competition" amongst the gametes produced by a single individual. 
As such, meiotic drive can only have an effect in heterozygotes.
Gametic competition on the other hand, is competition amongst all the gametes produced by one sex (e.g., pollen from many male plants).
Thus allele frequency changes can occur by gametic competition even if all individuals were homozygous (as long as there is a polymorphism).
We have extended our description of how `gametic competition' and `meiotic drive' are modelled in the model section (L106-116), which helps to clarify why they are different and under what circumstances each applies. 
}

### 2018-05-17 LaTeX/main_revised.tex
Sally's main text edits incorporated, not Figure changes after Figure 1

\end{flushleft}
% Please keep the abstract below 300 words
%currently 308 words
\section*{Abstract}
Sex determination is remarkably dynamic; many taxa display shifts in the location of sex-determining loci or the evolution of entirely new sex-determining systems. 
Predominant theories for why we observe such transitions generally conclude that novel sex-determining systems are favoured by selection if they equalise the sex ratio or increase linkage with a locus that experiences different selection in males vs.\ females. 
We use population genetic models to extend these theories in two ways: 
(1) We consider the dynamics of loci very tightly linked to the ancestral sex-determining loci, e.g., within the non-recombining region of the ancestral sex chromosomes. 
Variation at such loci can favour the spread of new sex-determining systems in which the heterogametic sex changes (XY to ZW or ZW to XY) and the new sex-determining region is less closely linked (or even unlinked) to the locus under selection. %, which is \textcolor{red}{not expected from previous theory}. 
(2) We consider selection upon haploid genotypes either during gametic competition (e.g., pollen competition) or meiosis (i.e., non-Mendelian segregation), which can cause the zygotic sex ratio to become biased. 
Haploid selection can drive transitions between sex-determining systems without requiring selection to act differently in diploid males vs.\ females. 
With haploid selection, we find that transitions between male and female heterogamety can evolve where linkage with the sex-determining locus is either strengthened or weakened. 
%Furthermore, we find that the selective strengths exerted by the sex-ratio and associations with selected alleles are often equivalent. 
Furthermore, we find that sex-ratio biases may increase or decrease with the spread of new sex chromosomes, which implies that transitions between sex-determining systems cannot be simply predicted by selection to equalise the sex ratio.
In fact, under many conditions, we find that transitions in sex determination are favoured equally strongly in cases where the sex ratio bias increases or decreases.
%Furthermore, we find that the selective forces to equalize the sex ratio are equally important to selective forces that generate skewed sex ratios when accounting for the spread of new sex chromosomes.
%This allows sex-ratio biases to increase or decrease with the spread of new sex chromosomes and implies that transitions between sex-determining systems cannot be simply predicted by selection to equalise the sex ratio. 
Overall, our models reveal that transitions between sex-determining systems, particularly transitions where the heterogametic sex changes, can be driven by loci in previously unpredicted genomic locations that experience selection during diploid and/or haploid phases.
%Our results predict conditions under which sex-determining systems are likely to be labile and draw connections with sex-ratio evolution.

% Please keep the Author Summary between 150 and 200 words
% Use first person. PLOS ONE authors please skip this step. 
% Author Summary not valid for PLOS ONE submissions.   
%currently 193 words
\section*{Author summary}
Systems of sex determination are strikingly diverse and labile in many clades. 
This poses the question: what drives transitions between sex-determining systems? 
Here, we use models to derive conditions under which new sex-determining systems spread. 
Prevailing views suggest that new sex-determining systems are favoured when they equalize the sex ratio and/or when they are more closely linked to genes that experience differential selection in males and females. 
Our models include selection upon haploid genotypes (meiotic drive or gametic competition), which biases the sex-ratio and occurs differently in male and female gametes. 
Surprisingly, we find the two forces (selection to equalize the sex ratio and the benefits of hitchhiking with driven alleles that distort the sex ratio) will often be equally strong, and thus neither is sufficient to explain the spread of new sex-determining systems in every case. 
We also find that new sex-determining alleles can spread despite being less closely linked to selected loci, driven by initially tight linkage or haploid selection. 
Our models therefore predict that loci in previously unexpected genomic locations and/or experiencing various types of selection (including haploid selection) can now be implicated as drivers of transitions between sex-determining systems.

@@

% Use "Eq" instead of "Equation" for equation citations.
\section*{Introduction}
%TRANSITIONS ARE COMMON
Animals and angiosperms exhibit extremely diverse sex-determining systems (reviewed in~\cite{Bull:1983vi,Charlesworth:2010it,Beukeboom:2014vb,Bachtrog:2014bx, Pennell2018}). 
Among species with genetic sex determination (GSD), some taxa have heterogametic males (XY) and homogametic females (XX), including %non-monotreme? I think monotremes have XY but tonnes of translocations, giving: X1,X2...X6 Y1
mammals and most dioecious plants~\cite{Ming:2011iy}; whereas other taxa have homogametic males (ZZ) and heterogametic females (ZW), including Lepidoptera and birds. 
Within several taxa, the chromosome that harbours the master sex-determining locus changes. 
For example, transitions of the master sex-determining locus between chromosomes or the evolution of new master sex-determining loci where the heterogametic sex does not change (hereafter `cis-GSD transitions') have occurred in Salmonids~\cite{Li:2011fm,Yano:2012di}, Diptera~\cite{Vicoso:2015hf}, and \textit{Oryzias}~\cite{Myosho:2012fv}. 
%Presentation at evolution found neo sex chromosome in birds, doesn't seem to be published or bioRxiv yet (consider pers com.), title: A previously unknown neo-sex chromosome mediates plumage divergence and speciation in hybridizing birds Jason Sardell; Elizabeth Cooper; J. Albert C. Uy. I wonder if this is actually a fusion event though?
In addition, many clades exhibit transitions between male and female heterogamety (XY$\leftrightarrow$ZW, hereafter `trans-GSD transitions'), including snakes~\cite{Gamble2017}, lizards~\cite{Ezaz:2009tk}, eight of 26 teleost fish families~\cite{Mank:2006bt}, true fruit flies (Tephritids,~\cite{Vicoso:2015hf}), amphibians~\cite{Hillis:1990gu}, the angiosperm genus \textit{Silene}~\cite{Slancarova:2013dq}, the angiosperm family \textit{Salicaceae}~\cite{Pucholt2015,Pucholt2017} and Coleoptera and Hemiptera (plate 2~\cite{Beukeboom:2014vb}).
Indeed, in some cases, both male and female heterogametic sex-determining systems can be found in the same species, as reported in houseflies~\cite{Macdonald1978}, midges~\cite{Thompson1971}, frogs~\cite{Ogata:2007jm}, cichlid fish~\cite{Ser:2010iq}, tilapia~\cite{Lee2004}, sea bass~\cite{Vandeputte2007}, and lab-strains of Zebrafish~\cite{Liew2012,Wilson2014}.
In addition, multiple transitions have occurred between genetic and environmental sex-determining systems (GSD$\leftrightarrow$ESD), e.g., in reptiles and fishes~\cite{Conover:1987in,Mank:2006bt,Pokorna:2009ui,Ezaz:2009tk,Pen:2010kk,Holleley:2015hc, Pennell2018}.
In sum, accumulating evidence indicates that transitions between sex-determining systems are common~\cite{Bachtrog:2014bx}.

%MAJOR MECHANISMS
%(1) sex-ratio selection
%Predominant theories accounting for transitions between GSD systems by selection involve sex-ratio selection or selective differences between the sexes (e.g., sexually antagonistic selection)~\cite{Blaser2012, Beukeboom:2014vb,vanDoorn2014re}. 
%\citet{Bull:1977wt} shows that direct selection on new sex determination loci can allow them to spread and switch between male and female heterogamety. 
It has been suggested that sex-ratio selection is a particularly dominant force in the evolution of sex determination because the sex ratio is directly determined by the sex-determining system (e.g., Bull, 1983, p 66-67 ~\cite{Bull:1983vi}; Buekeboom and Perrin, 2014, Chapter 7~\cite{Beukeboom:2014vb}). 
Classic `Fisherian' sex-ratio selection favours a 1:1 zygotic sex ratio when assuming that males and females are equally costly to produce~\cite{Fisher:1930wy,Charnov:1982wg}. 
This follows from the fact that, for an autosomal locus, half of the genetic material is inherited from a male and half from a female~\cite{West:2009we}. 
Thus, if the sex ratio is biased, an individual of the rarer sex will, on average, contribute more genetic material to the next generation. 
Selection therefore typically favours mutants that increase investment in the rarer sex.
%Because the sex ratio is directly determined by the sex-determining system, it has been suggested that sex-ratio selection is a particularly dominant force in the evolution of sex determination (e.g., Bull, 1983, p 66-67 ~\cite{Bull:1983vi}; Buekeboom and Perrin, 2014, Chapter 7~\cite{Beukeboom:2014vb}).

@@

%Tight linkage Model Feature and Result
One novel feature of the models developed here is that we explicitly consider the maintenance of genetic variation around the ancestral sex-determining locus (e.g., within the non-recombining region of a sex chromosome). %during trans-GSD transitions
%, when considering loci in very tight linkage with the ancestral sex-determining locus (e.g., within the non-recombining region of a sex chromosome), 
%\textcolor{red}{[vDK2007 does "explicitly" consider this as well (although on pg 15 of their SOM they say they will ignore equilibrium A, as it is stable under a "limited range of biologically relevant parameter combinations"; perhaps haploid selection changes this impression), but only for cis-GSD transitions. They find that if equil B is stable the neo-Y never spreads when r=0. vDK2010 does not "explicitly" calculate pX and pY, which are required to see if their equation 7 is positive or negative, but nowhere do they seem to say that weaker linkage \textit{can't} evolve given r initially small.]}
%we explicitly calculate ancestral polymorphisms. 
%Our analysis reveals that, under various forms of selection, the ancestral-X can carry suboptimal alleles for females and/or the ancestral-Y can carry female-beneficial alleles. 
Counterintuitively, when linkage is tight between the sex-determining locus and a selected locus, an allele good for females can be at higher frequency on the ancestral-Y than on the ancestral-X under a variety of forms of selection.
In addition, selection on ancestral-X chromosomes in males can prevent the X from becoming optimally specialised for female-beneficial alleles. 
These factors, in turn, can favour a new ZW sex-determining locus that has weaker linkage with loci under selection, which was not \textcolor{red}{apparent/discovered/considered} in previous theory \cite{vanDoorn:2010hu}. A similar argument applies to ZW$\rightarrow$XY transitions. 
Thus, we show that selected loci in very tight linkage with the ancestral GSD locus can favour trans-GSD transitions during which linkage associations are actually weakened.

%Haploid selection model feature and discussion of why needed
Most significantly, we include haploid selection (gametic competition or meiotic drive) in models describing cis-GSD, trans-GSD, and GSD to ESD transitions.
This poses an apparent evolutionary problem. 
On one hand, haploid selection is typically sex-limited in that it usually occurs among gametes produced by one sex only~\cite{Mulcahy:1996ha,JOSEPH:2004haa,Ubeda:2005gw,Lindholm:2016cw}. 
Therefore, one might expect new sex-determining systems to benefit from close linkage with haploid selected loci, as found for sex-differences in diploid selection~\cite{vanDoorn:2007eu,vanDoorn:2010hu, Muralidhar2018}. 
On the other hand, associations between sex-determining loci and haploid selected loci generate biased zygotic sex ratios, which should generally hinder the spread of new sex-determining systems.

>>>>

\end{flushleft}
% Please keep the abstract below 300 words
%currently 308 words
\section*{Abstract}
Sex determination is remarkably dynamic; many taxa display shifts in the location of sex-determining loci or the evolution of entirely new sex-determining systems. 
Predominant theories for why we observe such transitions generally conclude that novel sex-determining systems are favoured by selection if they equalise the sex ratio or increase linkage with a locus that experiences different selection in males vs.\ females. 
We use population genetic models to extend these theories in two ways: 
(1) We consider the dynamics of loci very tightly linked to the ancestral sex-determining loci, e.g., within the non-recombining region of the ancestral sex chromosomes. 
Variation at such loci can favour the spread of new sex-determining systems in which the heterogametic sex changes (XY to ZW or ZW to XY) and the new sex-determining region is less closely linked (or even unlinked) to the locus under selection. %, which is \textcolor{red}{not expected from previous theory}. 
(2) We consider selection upon haploid genotypes either during gametic competition (e.g., pollen competition) or meiosis (i.e., non-Mendelian segregation), which can cause the zygotic sex ratio to become biased. 
Haploid selection can drive transitions between sex-determining systems without requiring selection to act differently in diploid males vs.\ females. 
With haploid selection, we find that transitions between male and female heterogamety can evolve where linkage with the sex-determining locus is either strengthened or weakened. 
%Furthermore, we find that the selective strengths exerted by the sex-ratio and associations with selected alleles are often equivalent. 
Furthermore, we find that sex-ratio biases may increase or decrease with the spread of new sex chromosomes, which implies that transitions between sex-determining systems cannot be simply predicted by selection to equalise the sex ratio.
In fact, under many conditions, we find that transitions in sex determination are favoured equally strongly in cases where the sex ratio bias increases or decreases.
%Furthermore, we find that the selective forces to equalize the sex ratio are equally important to selective forces that generate skewed sex ratios when accounting for the spread of new sex chromosomes.
%This allows sex-ratio biases to increase or decrease with the spread of new sex chromosomes and implies that transitions between sex-determining systems cannot be simply predicted by selection to equalise the sex ratio. 
Overall, our models predict that sex determination systems should be highly dynamic, particularly when haploid selection is present, consistent with the evolutionary lability of this trait in many taxa.
%Overall, our models reveal that transitions between sex-determining systems, particularly transitions where the heterogametic sex changes, can be driven by loci in previously unpredicted genomic locations that experience selection during diploid and/or haploid phases.
%Our results predict conditions under which sex-determining systems are likely to be labile and draw connections with sex-ratio evolution.

% Please keep the Author Summary between 150 and 200 words
% Use first person. PLOS ONE authors please skip this step. 
% Author Summary not valid for PLOS ONE submissions.   
%currently 193 words
\section*{Author summary}
Systems of sex determination are strikingly diverse and labile in many clades. 
This poses the question: what drives transitions between sex-determining systems? 
Here, we use models to derive conditions under which new sex-determining systems spread. 
Prevailing views suggest that new sex-determining systems are favoured when they equalize the sex ratio and/or when they are more closely linked to genes that experience differential selection in males and females. 
Our models include selection upon haploid genotypes (meiotic drive or gametic competition), which biases the sex-ratio and occurs differently in male and female gametes. 
Surprisingly, we find the two forces (selection to equalize the sex ratio and the benefits of hitchhiking alongside driven alleles that distort the sex ratio) will often be equally strong, and thus neither is sufficient to explain the spread of new sex-determining systems in every case. 
We also find that new sex-determining alleles can spread despite being less closely linked to selected loci as long as initial linkage is tight or haploid selection is present.. 
Our models therefore predict that loci in previously unexpected genomic locations and/or experiencing various types of selection (including haploid selection) can now be implicated as drivers of transitions between sex-determining systems.

@@

% Use "Eq" instead of "Equation" for equation citations.
\section*{Introduction}
%TRANSITIONS ARE COMMON
Animals and angiosperms exhibit extremely diverse sex-determining systems (reviewed in~\cite{Bull:1983vi,Charlesworth:2010it,Beukeboom:2014vb,Bachtrog:2014bx, Pennell2018}). 
Among species with genetic sex determination (GSD), some taxa have heterogametic males (XY) and homogametic females (XX), including %non-monotreme? I think monotremes have XY but tonnes of translocations, giving: X1,X2...X6 Y1
mammals and most dioecious plants~\cite{Ming:2011iy}; whereas other taxa have homogametic males (ZZ) and heterogametic females (ZW), including Lepidoptera and birds. 
Within several taxa, the chromosome that harbours the master sex-determining locus changes, due either to translocation of the master sex-determining locus or to the evolution of a new master locus.  
During these transitions, the heterogametic sex can remain the same (hereafter `cis-GSD transitions') as in Salmonids~\cite{Li:2011fm,Yano:2012di}, Diptera~\cite{Vicoso:2015hf}, and \textit{Oryzias}~\cite{Myosho:2012fv}. 
%Presentation at evolution found neo sex chromosome in birds, doesn't seem to be published or bioRxiv yet (consider pers com.), title: A previously unknown neo-sex chromosome mediates plumage divergence and speciation in hybridizing birds Jason Sardell; Elizabeth Cooper; J. Albert C. Uy. I wonder if this is actually a fusion event though?
Alternatively, species can switch between male and female heterogamety (XY$\leftrightarrow$ZW, hereafter `trans-GSD transitions'), as in snakes~\cite{Gamble2017}, lizards~\cite{Ezaz:2009tk}, eight of 26 teleost fish families~\cite{Mank:2006bt}, true fruit flies (Tephritids,~\cite{Vicoso:2015hf}), amphibians~\cite{Hillis:1990gu}, the angiosperm genus \textit{Silene}~\cite{Slancarova:2013dq}, the angiosperm family \textit{Salicaceae}~\cite{Pucholt2015,Pucholt2017} and Coleoptera and Hemiptera (plate 2~\cite{Beukeboom:2014vb}).
Indeed, in some cases, both male and female heterogametic sex-determining systems can be found in the same species, as reported in houseflies~\cite{Macdonald1978}, midges~\cite{Thompson1971}, frogs~\cite{Ogata:2007jm}, cichlid fish~\cite{Ser:2010iq}, tilapia~\cite{Lee2004}, sea bass~\cite{Vandeputte2007}, and lab-strains of Zebrafish~\cite{Liew2012,Wilson2014}.
In addition, multiple transitions have occurred between genetic and environmental sex-determining systems (GSD$\leftrightarrow$ESD), e.g., in reptiles and fishes~\cite{Conover:1987in,Mank:2006bt,Pokorna:2009ui,Ezaz:2009tk,Pen:2010kk,Holleley:2015hc, Pennell2018}.
In sum, accumulating evidence indicates that transitions between sex-determining systems are common~\cite{Bachtrog:2014bx}.

%MAJOR MECHANISMS
%(1) sex-ratio selection
%Predominant theories accounting for transitions between GSD systems by selection involve sex-ratio selection or selective differences between the sexes (e.g., sexually antagonistic selection)~\cite{Blaser2012, Beukeboom:2014vb,vanDoorn2014re}. 
%\citet{Bull:1977wt} shows that direct selection on new sex determination loci can allow them to spread and switch between male and female heterogamety. 
It has been suggested that sex-ratio selection is a particularly dominant force in the evolution of sex determination (e.g., Bull, 1983, p 66-67 ~\cite{Bull:1983vi}; Buekeboom and Perrin, 2014, Chapter 7~\cite{Beukeboom:2014vb}). 
Classic `Fisherian' sex-ratio selection favours a 1:1 zygotic sex ratio when assuming that males and females are equally costly to produce~\cite{Fisher:1930wy,Charnov:1982wg}. 
This follows from the fact that, for an autosomal locus, half of the genetic material is inherited from a male and half from a female~\cite{West:2009we}. 
Thus, if the sex ratio is biased, an individual of the rarer sex will, on average, contribute more genetic material to the next generation. 
Selection therefore typically favours mutants that increase investment in the rarer sex, including new sex determination systems.
%Because the sex ratio is directly determined by the sex-determining system, it has been suggested that sex-ratio selection is a particularly dominant force in the evolution of sex determination (e.g., Bull, 1983, p 66-67 ~\cite{Bull:1983vi}; Buekeboom and Perrin, 2014, Chapter 7~\cite{Beukeboom:2014vb}).

@@

%Tight linkage Model Feature and Result
One novel feature of the models developed here is that we explicitly consider the maintenance of genetic variation around the ancestral sex-determining locus (e.g., within the non-recombining region of a sex chromosome). %during trans-GSD transitions
%, when considering loci in very tight linkage with the ancestral sex-determining locus (e.g., within the non-recombining region of a sex chromosome), 
%\textcolor{red}{[vDK2007 does "explicitly" consider this as well (although on pg 15 of their SOM they say they will ignore equilibrium A, as it is stable under a "limited range of biologically relevant parameter combinations"; perhaps haploid selection changes this impression), but only for cis-GSD transitions. They find that if equil B is stable the neo-Y never spreads when r=0. vDK2010 does not "explicitly" calculate pX and pY, which are required to see if their equation 7 is positive or negative, but nowhere do they seem to say that weaker linkage \textit{can't} evolve given r initially small.]}
%we explicitly calculate ancestral polymorphisms. 
%Our analysis reveals that, under various forms of selection, the ancestral-X can carry suboptimal alleles for females and/or the ancestral-Y can carry female-beneficial alleles. 
Counterintuitively, when linkage is tight between the sex-determining locus and a selected locus, an allele good for females can be at higher frequency on the ancestral-Y than on the ancestral-X under a variety of forms of selection.
In addition, selection on ancestral-X chromosomes in males can prevent the X from becoming optimally specialised for female-beneficial alleles. 
These factors, in turn, can favour a new ZW sex-determining locus that has weaker linkage with loci under selection, which was not revealed by previous theory \cite{vanDoorn:2010hu}. A similar argument applies to ZW$\rightarrow$XY transitions. 
Thus, we show that selected loci in very tight linkage with the ancestral GSD locus can favour trans-GSD transitions during which linkage associations are actually weakened.

%Haploid selection model feature and discussion of why needed
Most significantly, we include haploid selection (gametic competition or meiotic drive) in models describing cis-GSD, trans-GSD, and GSD to ESD transitions.
This poses an apparent evolutionary problem. 
On one hand, haploid selection is typically sex-limited in that it usually occurs among gametes produced by one sex only~\cite{Mulcahy:1996ha,JOSEPH:2004haa,Ubeda:2005gw,Lindholm:2016cw}. 
Therefore, one might expect new sex-determining systems to benefit from close linkage with haploid selected loci, as found for loci that experience diploid-sex-differences in selection~\cite{vanDoorn:2007eu,vanDoorn:2010hu, Muralidhar2018}. 
On the other hand, associations between sex-determining loci and haploid selected loci generate biased zygotic sex ratios, which should generally hinder the spread of new sex-determining systems.

%Previous models with haploid selection
Two previous studies have considered the spread of GSD with sex-limited meiotic drive~\cite{Kozielska:2010vm,Ubeda:2015fx} under a limited number of possible genetic architectures and diploid selective regimes. 
Ubeda et al. (2015)~\cite{Ubeda:2015fx} considered ancestral-ESD (with no sex-ratio bias) and numerically showed that new GSD alleles can spread if they arise in linkage with meiotic drive loci. 
For example, a masculinizing allele spreads in association with an allele that is favoured during male meiosis, causing sex ratios to become male-biased.
This suggests that the benefits of associating with driving alleles can overwhelm selection to balance the sex ratio. %, at least when new sex-determining alleles are rare. 
However, Kozielska et al. (2010)~\cite{Kozielska:2010vm} considered an ancestral GSD system that is perfectly linked to a meiotic driver and therefore exhibiting an ancestral sex ratio bias. 
They found that a new, completely unlinked, GSD system can spread if it generates the rarer sex, creating a balanced sex ratio. 
This suggests that Fisherian sex-ratio selection can overwhelm the benefits of being associated with driving alleles. 
It is thus currently unclear when haploid selection favors increased versus decreased linkage between haploid selected loci and a new sex-determining locus. 
In addition, because the sex ratio is determined by linkage between haploid selected loci and the sex-determining locus, it is also unclear when Fisherian sex-ratio selection is the most important driver of transitions between sex-determining systems.
